Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-04-05 pkg/disttree/inst/draft/draft.Rnw
final touches incorporating Torsten's and Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2127 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 44a48df3d37767b050b25054e7f90ee80823d816

\documentclass[nojss]{jss}

%\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
%\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}
%\VignettePackage{disttree}

@@

\hyphenation{Qua-dra-tic}

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

@@

transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)

@@

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameter(s) of the distribution are linked to regressors. In many classical
models this only captures the location/expectation of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting and, more generally, the GAMLSS framework
allows to establish generalized additive models for location, scale, and shape
with smooth linear or nonlinear effects.
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous locations in a mountainous
region (Tyrol, Austria) based on a very large number of numerical weather
prediction quantities. It is shown that distributional random forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, distributional regression}

@@

While in many applications conditional mean regression models have been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
instead there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. As an example, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear 
deterministic models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWPs. Thus, the NGR
considers both the mean as well as the uncertainty of the ensemble predictions
to obtain probabilistic forecasts calibrated to a particular site.

@@

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{fig:tree}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional trees} split the data into more homogeneous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
  \item \emph{Distributional forests} utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects and interactions where it would be 
challenging to specify a distributional regression model like GAMLSS.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=4>>=
par(mar = c(4, 4, 3, 1.5))
plot(x = x, y = y1, type = "l", col = "red", 
     main = paste0("July ", pday), ylab = "Density", 
     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
     xlim = c(-1.5,8))

lines(x = x, y = y2, type = "l", col = "blue")
lines(x = x, y = y3, type = "l", col = "darkgreen")
lines(x = x, y = y4, type = "l", col = "purple")
legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8)

# plot point mass
lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = "red", type = "l", lwd = 1)
lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = "blue", type = "l", lwd = 1)
lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = "darkgreen", type = "l", lwd = 1)
lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = "purple", type = "l", lwd = 1)

points(x = pm1[1], y = pm1[2], col = "red", pch = 19)
points(x = pm2[1], y = pm2[2], col = "blue", pch = 19)
points(x = pm3[1], y = pm3[2], col = "darkgreen", pch = 19)
points(x = pm4[1], y = pm4[2], col = "purple", pch = 19)

# plot predictions
points(x = pred1[1], y = pred1[2], col = "red", pch = 4)
points(x = pred2[1], y = pred2[2], col = "blue", pch = 4)
points(x = pred3[1], y = pred3[2], col = "darkgreen", pch = 4)
points(x = pred4[1], y = pred4[2], col = "purple", pch = 4)

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2)

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = "red", cex = 0.8)
text(x = -0.8, y = lh2, labels = "2010", col = "blue", cex = 0.8)
text(x = -0.8, y = lh3, labels = "2011", col = "darkgreen", cex = 0.8)
text(x = -0.8, y = lh4, labels = "2012", col = "purple", cex = 0.8)
@
\caption{\label{fig:axams-dist}Predictions of total precipitation at station Axams for July 24 in 2009, 2010, 2011 and 2012 by a distributional forest learned on data from 1985--2008. Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
\end{figure}

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly non-linear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a zero left-censored Gaussian
distribution is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise models yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as training data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
It is shown that they perform at least on par and sometimes clearly better than
three alternative zero-censored Gaussian models: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as a technique for variable selection among all
80~available regressors.

@@

The general distributional notation is exemplified in all three steps for the
zero-censored Gaussian distribution. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$. Based on the
corresponding probability density function $f(Y; \bm{\theta})$ the log-likelihood
function is defined $\ell(\bm{\theta}; Y) = \log\{f(Y; \bm{\theta})\}$. The GAMLSS
framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
distributional families with parameterizations corresponding to location, scale,
and shape. Furthermore, censoring and/or truncation of these distributions
can be incorporated in the usual straightforward way
\citep[see e.g.,][Chapter~7.2]{Long:1997}.

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), zero-censored Gaussian distribution with mean/location parameter
$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    %\log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    %\log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
    \log\left\{\sigma^{-1} \cdot \phi\left((Y - \mu) \cdot \sigma^{-1}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left((-\mu) \cdto \sigma^{-1}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$

@@

With the specification of the distribution family and its log-likelihood 
function the task of fitting a distributional model turns into the task 
of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
maximum likelihood (ML) based on the learning sample
$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
$$ \label{eq:mle}
\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
$$
Equivalently, this can be defined based on the corresponding first-order
conditions
$$ \label{eq:foc}
\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
$$
where $s(\bm{\hat \theta}, y_i)$ is the associated score function
$$ \label{eq:score}
s(\bm{\hat{\theta}}, y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i) \mid_{\bm{\theta} = \bm{\hat{\theta}}}.
$$ 
The latter is subsequently employed as a general goodness-of-fit measure
to assess how well the distribution with parameters $\bm{\theta}$ fits
one individual observation $y_i$.

@@

Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
for reasonably representing the response distribution and covariates
$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
employed to capture differences in the distribution parameters $\bm{\theta}$.
In weather forecasting, these covariates typically include the output
from numerical weather prediction systems and/or lagged observations.

To incorporate the covariates into the distributional model, GAMLSS considers
them as regressors in additive predictors
$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$.
Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
nonlinear effects, spatial effects, random coefficients, or interaction surfaces
\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
additive terms and their functional forms in advance which can be challenging in
practice, especially if the number of covariates $m$ is large.

Regression trees generally take a different approach for automatically
including covariates in a data-driven way and allowing for abrupt
changes, non-additive effects, and interactions. In the context of
distributional models the goal is to partition the covariate space
$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
distributional model for the response $Y$ can be found with segment-specific
parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
($b = 1, \dots, B$) partition the covariate space
$$ \label{eq:partition}
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
$$
and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
response $Y$ in each segment.

@@

While the simple recursive structure of a tree model is easy to visualize and
interpret, the abrupt changes are often too rough, instable, and impose steps
on the model even if the true underlying effect is smooth. Hence, ensemble
methods such as bagging or random forests \citep{Breiman:2001} typically smooth
the effects, stabilize the model, and improve predictive performance.

While classical random forests \citep{Breiman:2001} grow ensembles of trees
that pick up changes in the location of the response across the covariates,
\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
pick up changes in the ``direction'' of any distribution parameter by
considering the full score vector for choosing splitting variables and split
points. Each of the distributional trees is grown on a different data set
obtained through bootstrap sampling (or subsampling) and in each node only a
random subset of the covariates $\bold{Z}$ is considered. As usual in random
forests, this reduces the correlation among the trees and stabilizes the
variance of the model.

To obtain probabilistic predictions from a distributional forest, it still
needs to be specified how to compute the parameter estimates
$\bm{\hat \theta}(\bold{z})$ for a (potentially new) set of covariates $\bold{z}$.
Following \cite{Hothorn+Zeileis:2017} we interpret random forests as adaptive
local likelihood estimators using the averaged ``nearest neighbor weights''
\citep{Lin+Jeon:2006} from the $T$~trees in the forest
$$
w^{\text{forest}}_i(\bold{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} \mathbf{1}((\bold{z}_i \in \mathcal{B}^t_b) \land (\bold{z} \in \mathcal{B}^t_b))
$$
Thus, these $w^{\text{forest}}_i(\bold{z}) \in [0, 1]$ whereas
$w^{\text{tree}}_i(\bold{z}) \in \{0, 1\}$. Hence, weights cannot only be $0$
or $1$ but change more smoothly, giving high weight to those observations $i$
from the learning sample that co-occur in the same segment $\mathcal{B}_b^t$
as the new observation $\bold{z}$ for many of the trees $t = 1, \dots, T$.
Consequently, the parameter estimates may, in principle, change for every
observation and can be obtained by
$$
\bm{\hat \theta}(\bold{z}) =  \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{forest}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
$$
In summary, this yields a parametric distributional regression model 
(through the score-based approach) that can capture both abrupt effects and
high-order interactions (through the trees) and smooth effects (through
the forest).

>>>>

\documentclass[nojss]{jss}
\shortcites{Hamill+Bates+Whitaker:2013}

%\VignetteIndexEntry{disttree: Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain}
%\VignetteDepends{disttree, crch, countreg, gamlss, gamlss.cens}
%\VignetteKeywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}
%\VignettePackage{disttree}

@@

@@

transpgray <- rgb(0.190,0.190,0.190, alpha = 0.2)
@

@@

\Abstract{
To obtain a probabilistic model for a dependent variable based on some set of
explanatory variables, a distributional approach is often adopted where the
parameters of the distribution are linked to regressors. In many classical
models this only captures the location of the distribution but
over the last decade there has been increasing interest in distributional
regression approaches modeling all parameters including location, scale, and
shape. Notably, so-called non-homogenous Gaussian regression (NGR) models both
mean and variance of a Gaussian response and is particularly
popular in weather forecasting. More generally, the GAMLSS framework
allows to establish generalized additive models for location, scale, and shape
with smooth linear or nonlinear effects.
%
However, when variable selection is required and/or there are
non-smooth dependencies or interactions (especially unknown or of high-order),
it is challenging to establish a good GAMLSS. A natural alternative in these
situations would be the application of regression trees or random forests but,
so far, no general distributional framework is available for these. Therefore,
a framework for distributional regression trees and forests is proposed that blends
regression trees and random forests with classical distributions from the GAMLSS
framework as well as their censored or truncated counterparts.
%
To illustrate these novel approaches in practice, they are employed to obtain
probabilistic precipitation forecasts at numerous sites in a mountainous
region (Tyrol, Austria) based on a large number of numerical weather
prediction quantities. It is shown that the novel distributional regression forests
automatically select variables and interactions, performing on par or often even
better than GAMLSS specified either through prior meteorological knowledge or a
computationally more demanding boosting approach.
}
\Keywords{parametric models, regression trees, random forests, recursive partitioning, probabilistic forecasting, GAMLSS}

@@

While in many applications conditional mean regression models have been receiving
the most attention, there has been a paradigm shift over the last decade towards
distributional regression models. An important reason for this is that in many
fields forecasts of the mean are not the only (or not even the main) concern but
instead there is an increasing interest in probabilistic forecasts. Quantities of
interest typically include exceedence probabilities for certain thresholds of the
response or quantiles of the response distribution. Specifically, consider
weather forecasting where there is less interest in the mean amount of
precipitation on the next day. Instead, the probability of rain vs.\ no rain
is typically more relevant or, in some situations, a prediction interval of
expected precipitation (say from the expected 10\% to 90\% quantile). Similar
considerations apply for other meteorological quantities and hence attention
in the weather forecasting literature has been shifting from classical linear 
deterministic models \citep{Glahn+Lowry:1972} towards probabilistic models such as the
non-homogeneous Gaussian regression (NGR) of \cite{Gneiting+Raftery+Westveld:2005}.
The NGR typically describes the mean of some meteorological response variable through
the average of the corresponding quantity from an ensemble of physically-based
numerical weather predictions (NWPs). Similarly, the variance of the response
is captured through the variance of the ensemble of NWPs. Thus, the NGR
considers both the mean as well as the uncertainty of the ensemble predictions
to obtain probabilistic forecasts calibrated to a particular site.

@@

In summary, the models discussed so far provide a broad and powerful toolset for
parametric distributional fits depending on a specified set of additive linear or
smooth nonlinear terms. A rather different approach to capturing the dependence on
covariates are tree-based models.
\begin{itemize}
  \item \emph{Regression trees} (\citealt{Breiman+Friedman+Stone:1984}) recursively
    split the data into more homogeneous subgroups and can thus capture abrupt shifts
    (Figure~\ref{fig:tree}, top left) and approximate nonlinear functions.
    Furthermore, trees automatically carry out a forward selection of covariates and
    their interactions. 
  \item \emph{Random forests} (\citealt{Breiman:2001}) average the predictions of
    an ensemble of trees fitted to resampled versions of the learning data. This
    stabilizes the recursive partitions from individual trees and hence better
    approximates smooth functions (Figure~\ref{fig:tree}, top middle)
\end{itemize}
While classical regression trees and random forests only model the mean of the response
we propose to follow the ideas from GAMLSS modeling -- as outlined
in Figure~\ref{fig:gam} -- and combine tree-based methods with parametric distributional models,
yielding two novel techniques:
\begin{itemize}
  \item \emph{Distributional regression trees} (for short: \emph{distributional trees}) split the data into more homogeneous groups with
    respect to a parametric distribution, thus capturing changes in any distribution
    parameter like location, scale, or shape (Figure~\ref{fig:tree}, bottom left)
  \item \emph{Distributional regression forests} (for short: \emph{distributional forests}) utilize an ensemble of distributional trees
    for obtaining stabilized and smoothed parametric predictions (Figure~\ref{fig:tree},
    top right).
\end{itemize}
In the following, particular focus is given to distributional forests as a method for obtaining
probabilistic forecasts by leveraging the strengths of random forests: the ability to
capture both smooth and abruptly changing functions along with simultaneous selection of
variables and possibly complex interactions. Thus, these properties make the method particularly
appealing in case of many covariates with unknown effects and interactions where it would be 
challenging to specify a distributional regression model like GAMLSS.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.75\textwidth}
<<rain_Axams_pred, fig=TRUE, echo=FALSE, width=6.5, height=3.8>>=
par(mar = c(3.8, 4, 2.3, 1.5))
plot(x = x, y = y1, type = "l", col = pal[1], lwd = 1.3, 
     main = paste0("July ", pday), ylab = "Density", 
     xlab = expression(Total~precipitation~"["~mm^(1/1.6)~"/"~"24h"~"]"),
     ylim = c(0,max(y1, y2, y3, y4, pm1, pm2, pm3, pm4) + 0.01),
     xlim = c(-1.5,8))

lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)

# plot point mass
lines(x = c(pm1[1], pm1[1]), y = c(pm1[2], 0), col = pal[1], type = "l", lwd = 1.3)
lines(x = c(pm2[1], pm2[1]), y = c(pm2[2], 0), col = pal[3], type = "l", lwd = 1.3)
lines(x = c(pm3[1], pm3[1]), y = c(pm3[2], 0), col = pal[2], type = "l", lwd = 1.3)
lines(x = c(pm4[1], pm4[1]), y = c(pm4[2], 0), col = pal[4], type = "l", lwd = 1.3)

points(x = pm1[1], y = pm1[2], col = pal[1], pch = 19)
points(x = pm2[1], y = pm2[2], col = pal[3], pch = 19)
points(x = pm3[1], y = pm3[2], col = pal[2], pch = 19)
points(x = pm4[1], y = pm4[2], col = pal[4], pch = 19)

# plot predictions
points(x = pred1[1], y = pred1[2], col = pal[1], pch = 4)
points(x = pred2[1], y = pred2[2], col = pal[3], pch = 4)
points(x = pred3[1], y = pred3[2], col = pal[2], pch = 4)
points(x = pred4[1], y = pred4[2], col = pal[4], pch = 4)

lines(x = c(pred1[1], pred1[1]), y = c(pred1[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred2[1], pred2[1]), y = c(pred2[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred3[1], pred3[1]), y = c(pred3[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)
lines(x = c(pred4[1], pred4[1]), y = c(pred4[2], 0), col = "darkgray", type = "l", lty = 2, lwd = 1.3)

# add labels
text(x = -0.8, y = lh1, labels = "2009", col = pal[1], cex = 0.8)
text(x = -0.8, y = lh2, labels = "2010", col = pal[3], cex = 0.8)
text(x = -0.8, y = lh3, labels = "2011", col = pal[2], cex = 0.8)
text(x = -0.8, y = lh4, labels = "2012", col = pal[4], cex = 0.8)
@
\caption{\label{fig:axams-dist}Total precipitation predictions by a distributional forest
at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
Observations are left-censored at 0. The corresponding predicted point mass is shown at the censoring point (0).}
\end{figure}

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution, left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no expert
knowledge for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

The general distributional notation is exemplified in all three steps for the
Gaussian distribution, left-censored at zero. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

A distributional model $\mathcal{D}(Y, \bm{\theta})$ is considered for the response
variable $Y \in \mathcal{Y}$ using the distributional family $\mathcal{D}$ with
\textit{k}-dimensional parameter vector $\bm{\theta} \in \bold{\Theta}$ and
corresponding log-likelihood function $\ell(\bm{\theta}; Y)$. The GAMLSS
framework \citep{Rigby+Stasinopoulos:2005} provides a wide range of such
distributional families with parameterizations corresponding to location, scale,
and shape. Furthermore, censoring and/or truncation of these distributions
can be incorporated in the usual straightforward way
\citep[see e.g.,][Chapter~7.2]{Long:1997}.

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$

@@

With the specification of the distribution family and its log-likelihood 
function the task of fitting a distributional model turns into the task 
of estimating the distribution parameter~$\bm{\theta}$. This is commonly done by
maximum likelihood (ML) based on the learning sample with observations
$\{y_i\}_{i = 1, \dots, n}$ of the response variable $Y$. The maximum
likelihood estimator (MLE) $\bm{\hat \theta}$ is given by
$$ \label{eq:mle}
\bm{\hat \theta} = \argmax_{\bm{\theta} \in \bold{\Theta}} \sum_{i=1}^n \ell(\bm{\theta}; y_i).
$$
Equivalently, this can be defined based on the corresponding first-order
conditions
$$ \label{eq:foc}
\sum_{i = 1}^n s(\bm{\hat{\theta}}, y_i) = 0,
$$
where $s(\bm{\theta}; y_i)$ is the associated score function
$$ \label{eq:score}
s(\bm{\theta}; y_i) = \frac{\partial \ell}{\partial \bm{\theta}}(\bm{\theta}; y_i).
$$ 
The latter is subsequently employed as a general goodness-of-fit measure
to assess how well the distribution with parameters $\bm{\theta}$ fits
one individual observation $y_i$.

@@

Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
for reasonably representing the response distribution and covariates
$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
employed to capture differences in the distribution parameters $\bm{\theta}$.
In weather forecasting, these covariates typically include the output
from numerical weather prediction systems and/or lagged meteorological observations.

To incorporate the covariates into the distributional model, 
they are considered as regressors in additive predictors
$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
in GAMLSS.
Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
nonlinear effects, spatial effects, random coefficients, or interaction surfaces
\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
additive terms and their functional forms in advance which can be challenging in
practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.

Regression trees generally take a different approach for automatically
including covariates in a data-driven way and allowing for abrupt
changes, nonlinear and non-additive effects, and interactions. In the context of
distributional models the goal is to partition the covariate space
$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
distributional model for the response $Y$ can be found with segment-specific
parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
($b = 1, \dots, B$) partition the covariate space
$$ \label{eq:partition}
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
$$
and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
response $Y$ in each segment.

@@

While the simple recursive structure of a tree model is easy to visualize and
interpret, the abrupt changes are often too rough, instable, and impose steps
on the model even if the true underlying effect is smooth. Hence, ensemble
methods such as bagging or random forests \citep{Breiman:2001} are typically
applied to smooth the effects, stabilize the model, and improve predictive performance.

While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
that pick up changes in the location of the response across the covariates,
\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
pick up changes in the ``direction'' of any distribution parameter by
considering the full score vector for choosing splitting variables and split
points. Each of the distributional trees is grown on a different data set
obtained through bootstrap sampling (or subsampling) and in each node only a
random subset of the covariates $\bold{Z}$ is considered. As usual in random
forests, this reduces the correlation among the trees and stabilizes the
variance of the model.
